Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-13 documents/ms.tex
small changes to the text

\newcommand{\dfmcomment}[1]{{\color{red}#1}}
\newcommand{\tdmcomment}[1]{{\color{blue}#1}}
\newcommand{\vrcomment}[1]{{\color{magenta}#1}}

@@

\begin{abstract}
The light curves of spotted, rotating stars often vary in a non-sinusoidal and
quasi-periodic fashion---spots move on the stellar surface and have finite
lifetimes, causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid therefore cannot accurately model a
rotationally modulated stellar light curve.
Physical models of the stellar surface also have many drawbacks preventing
    effective inference, such as highly degenerate or high-dimensional
    parameter spaces.
% \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In this work, we introduce an appropriate {\it effective} model: a Gaussian
    Process (GP) with a quasi-periodic covariance kernel function.
Using this highly flexible model, we are able to sample from the posterior
    Probability Density Function (PDF) of the periodic parameter,
    marginalising over the other kernel hyperparameters using a Markov Chain
    Monte Carlo approach.
To test the effectiveness of this method, we infer rotation periods from
\naigrain\ simulated stellar light curves, demonstrating that our results
    are more accurate than both a sine-fitting periodogram and an
    AutoCorrelation Function (ACF) method.
We also demonstrate that it works effectively on real data, by
inferring rotation periods for \nkois\ \Kepler\ stars.
Because this method delivers posteriors PDFs it will enable hierarchical
    studies involving stellar rotation, particularly those involving
    population modelling, such as inferring stellar ages, obliquities in
    exoplanet systems, or characterising star-planet interactions.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

Standard methods to measure rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[e.g.][]{Reinhold2013}, Auto-Correlation Functions (ACFs),
\citep{Mcquillan13b} and wavelet transforms \citep{Garcia2014}.
The precisions of the LS periodogram and wavelet methods are limited by the
suitability of the model choice: a sinusoid for the LS periodogram,
and a choice of mother wavelet, assumed to describe the data
over a range of transpositions \citep[see, \eg][]{Carter2010},
for the wavelet method.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, as long as the signal is approximately periodic the ACF will
display a peak at the rotation period, no matter its shape.
A drawback of the ACF method, however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.}, which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for most ground-based measurements, such
as the future Large Synoptic Survey Telescope (\LSST).
An ACF is also an operation performed on the data rather than a generative
model of the data, and so is not inherently probabilistic.
This means that the effects of the observational uncertainties cannot be
formally propagated to constraints on the rotation period.
% Many rotation periods in the literature have been inferred by measuring the
% position of the first peak in an ACF, however this approach can be dangerous.
% The exponential decay in correlation can shift the peak position short-wards
% of its true value, leading to an underestimate of the period.
% We return to this point in section \textsection \ref{sec:discussion}.

In this work, we introduce an {\it effective} model for rotationally
modulated stellar light curves
which captures the salient behaviour but is not
physically motivated---although some parameters may indeed be
{\it interpreted} as physical ones.
An ideal effective model should have a small number of non-degenerate parameters
and be flexible enough to perfectly capture non-sinusoidal and QP behaviour.
A Gaussian process (GP) model fulfills these requirements. We thus use a GP
as the generative model at the core of a method to probabilistically
infer accurate and precise stellar rotation periods.  This enables us to
estimate the posterior PDF of the rotation period, thereby producing a
justified estimate of its uncertainty while simultaneously marginalizing over
parameters which model correlated noise.

@@

GP models parameterise the covariance between data points by means of a
parametric kernel function that models the autocorrelation behaviour of the
signal and defines the form of the covariance matrix.
As a qualitative demonstration, we present the time-series in Figure
\ref{fig:GP_example}: the \kepler\ light curve of KIC \kepexample.
This is a relatively active star that rotates once every $\sim$
\kepexampleperiod\ days, with stochastic variability typical of \kepler\ FGK
stars.
Clearly, data points in this light curve are correlated.
Points close together in time are tightly correlated, and points more
widely separated are loosely correlated.
A GP models this variation in correlation with the separation between data
points; that is, it models the {\it covariance structure} rather than the data
directly.
This lends GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas modelling the time series itself
might require much more complexity.
Figure \ref{fig:GP_example} demonstrates how a GP model fits the light curve of
KIC \kepexample.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

A range of kernel functions can describe stellar variability.
For example, the most commonly used kernel function, the `Squared
Exponential' (SE), defined as follows, could adequately fit the
KIC \kepexample\ light curve:
\begin{equation}
\label{eq:SE}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
Here $A>0$ is the amplitude of covariance, $l$ is the length scale of
exponential decay, and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple, with
just two parameters, $A$ and $l$.
If $l$ is large, two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable, making it possible to model a data set
and its derivatives simultaneously.
% \citep[see \eg][]{Rajpaul2015}
However, The SE kernel function does not well describe the covariance
in stellar light curves, nor is it {\it useful} for the problem of
rotation period inference because it does not capture periodic behaviour.
Inferring rotation periods thus requires a periodic kernel
function.
For this reason, we use the `Quasi-Periodic' kernel.
\citet{Rasmussen2005} model QP variability in CO$_2$ concentration on the
summit of the Mauna Loa volcano in Hawaii \citep[data from][]{Keeling2004}
using a kernel which is the product of a periodic and a SE kernel: the QP
kernel.
This kernel is defined as
\begin{equation}
\label{eq:QP}
k_{i,j} = A \exp \left[-\frac{(x_i - x_j)^2}{2l^2} -
    \Gamma^2 \sin^2\left(\frac{\pi(x_i - x_j)}{P}\right) \right] + \sigma^2
    \delta_{ij}.
\end{equation}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star, and $\Gamma$
controls the amplitude of the $\sin^2$ term.
If $\Gamma$ is very large, only points almost exactly one period away are
tightly correlated and points that are slightly more or less than one period
away are very loosely correlated.
If $\Gamma$ is small, points separated by one period are tightly
correlated, and points separated by slightly more or less are still highly
correlated, although less so.
In other words, large values of $\Gamma$ lead to periodic variations with
increasingly complex harmonic content.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while also allowing
points separated by half a period to be weakly correlated.
We also use an extra parameter $\sigma$ which is an additional white noise
term added to the diagonal elements of the covariance matrix.
This can be interpreted to represent underestimation of observational
uncertainties~---~if the uncertainties reported on the data are too small, it
will be non-zero~---~or it can capture any remaining ``jitter'' or residuals
that are not captured by the effective GP model.
We use this QP kernel function to produce the GP model
that fits the \Kepler\ light curve in in Figure \ref{fig:GP_example}.
There are many ways to construct a periodic kernel and there are several
non-periodic kernel choices, other than the SE kernel that could be combined
with a periodic kernel in order to produce a QP kernel.
The kernel presented above reproduces the behaviour of stellar light curves
but it is not necessarily the only kernel choice that can do so.
We do not attempt to test any other choices in this paper, noting only that
this kernel function provides an adequeate model and leaving formal
comparisons with other kernel function choices to a future publication.

To infer a stellar rotation period $P$ from a light curve, we fit this
QP-kernel GP model to the data.  As with any model-fitting exercise, the
likelihood
of the model could be maximized to find the maximum-likelihood value for $P$.
In this study, however, we explore the full posterior PDFs using a Markov
Chain Monte Carlo (MCMC) procedure.  While
this approach comes at a computational cost, such posterior exploration
importantly provides a justified uncertainty estimate.

@@

This paper is laid out as follows.
In \textsection \ref{sec:simulations} we describe the simulated light curves
used to test the three rotation period recovery methods (ACF, LS periodogram
and GP).
The GP method is described in \textsection \ref{sec:method}.
Its performance is demonstrated and compared with literature methods in
in \textsection \ref{sec:perf_and_comp}.
In \textsection \ref{sec:kepler} we apply our method to real \kepler\ data,
and the results are discussed in \textsection \ref{sec:discussion}.

@@

The flexibility of this GP model allows for posterior multimodality and
``over-fitting''--like behavior.
For example, if $l$ is small, the non-periodic factor in the covariance
kernel may dominate, allowing for a good fit to the data without
requiring any periodic covariance structure---even if clear periodic
structure is present.
Additionally, if $\Gamma$ is too large, the GP model becomes extremely
flexible and can fit the data without varying the period.
Managing this flexibility to reliably retrieve the period parameter requires
imposing informative priors on the non-period GP parameters.
% While there are no strictly {\it a priori} reasons to believe
% that any of the hyperparameters should take values in any specific range,
% we have found in this work that reasonable priors for any particular
% dataset can be determined by experimentation.
% We note this is a somewhat subjective art, with a goal of
% not allowing the GP to become so flexible that it does not require
% periodicity to fit the data even in the presence of clear periodicity.
In particular, we find it necessary to avoid large values of
$A$ and $\Gamma$, and small values of $l$; though the exact details of what
works well may differ among datasets (see \ref{sec:} and
\ref{sec:} for two different examples of hyperparameter priors).
More physically motivated priors are discussed in \textsection
\ref{sec:discussion}.

@@

While using an uninformed prior on period (e.g., log-flat) often suffices,
we also develop a method to construct an informed period prior, based on the
autocorrelation function (ACF).
The ACF has proven to be very useful for measuring stellar rotation periods
\citep{Mcquillan2012, Mcquillan13b, Mcquillan2014}; however, the
method has several shortcomings,
most notably the inability to deliver uncertainties, but also
the necessity of several heuristic choices,
such as a timescale on which to smooth the ACF,
how to define a peak, whether the first or second peak
gets selected, and what constitutes a secure detection.
While this paper presents a rotation period inference method
that avoids these shortcomings,
it seems prudent to still use information available from the ACF.
We thus use the ACF to define a prior on period,
which can help the posterior sampling converge on the true period.

@@

\item{We assign a quality metric $Q_i$ to each of these candidate
periods, as follows.  First, we model the ACF as a
damped oscillator with fixed period $T_i$:
\begin{equation}
y = A e^{-t/\tau} \cos{\frac{2\pi t}{T_i} },
\end{equation}
where and $t$ is the lag time,
and find the best-fitting parameters $A_i$ and $\tau_i$ by a non-linear
least squares minimization procedure.  We then define the
following heuristic quality metric:
\begin{equation}
\label{eq:quality}
Q_i = \left(\frac{\tau_i}{T_i}\right) \left(\frac{N_i h_i}{R_i}\right),
\end{equation}
where $h_i$ is the height of the ACF peak at $T_i$,
$N_i$ is the length of the lag vector in the ACF (directly proportional
to the maximum allowed period $P_i$),
and $R_i$ is the sum of squared residuals between the
damped oscillator model and the actual ACF data.  The idea behind this
quality metric to give a candidate period a higher score if

@@

We initialize 500 walkers with random samples from the prior and use a
weighted mixture of the KDE, DE, and snooker proposals with weights of 0.4,
0.4, and 0.2 respectively.
We run \vrcomment{50} steps of the sampler at a time, checking for convergence
after each iteration, up to a maximum of \vrcomment{50} iterations.
We declare convergence if the total effective chain length is at least
8$\times$ the maximum autocorrelation time.
When convergence is achieved, we discard the first two autocorrelation lengths
in the chain as a burn-in, and randomly choose 5000 samples as representative
of the posterior.
This fitting process takes several hours for a typical simulated light curve,
though in some cases it can take 12 hours or longer to converge.

@@

In order to benchmark this new rotation period recovery method, we apply
it to a set of simulated light curves
and compare to the performance of established literature methods.
Section \ref{sec:simulations} describes the simulated data we use;
Section \ref{sec:performance} demonstrates the performance of the
QP-GP method; and Section \ref{sec:comparison} compares to the performance
of the Lomb-Scargle periodogram and autocorrelation function methods.

@@

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarize our results compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
To assess the performance of the QP-GP and other period recovery
methods, we compute Median Absolute Deviations (MADs) of the results, relative
to the input periods.
We also compute the Median Relative Deviation (MRD), as a percentage.
These metrics are presented for the three different methods tested in this
paper in table \ref{tab:MADs}.
The informative and uninformative prior versions of the GP method have MRDs of
\percentgpMAD \% and \percentgpMADnp \% respectively.
The marginal posterior distributions of the QP kernel hyperparameters, for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{figures/compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{figures/comparison_acfprior_02_13.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique.
Since the posterior PDFs of rotation periods are often non-Gaussian,
the points plotted here are the highest likelihood samples.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
The ACF-informed prior on rotation period used to generate these results is
    described in the appendix.
The MAD of these results is 0.38 in log-space, \ie\ 1.46 days.
    }
\label{fig:compare_mcmc_acfprior}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/comparison_noprior_02_13.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique with no .
Since the posterior PDFs of rotation periods are often non-Gaussian,
the points plotted here are the highest likelihood samples.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
An uninformative prior, flat in the natural log of the rotation period between
    0.5 and 100 days was used to generate these results.
The MAD of these results is 0.48 in log-space, \ie\ 1.61 days.
    }
\label{fig:compare_mcmc_noprior}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/corner_plot.pdf}
\caption{Marginal posterior PDFs of the QP GP model parameters, fit to the
    light curve in figure \ref{fig:demo_lc}.
    The blue line in the period panel shows the injected period.
    This figure was made using \texttt{corner.py} \citep{Corner}.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

@@

We apply the ACF method to the sample of \naigrain\ simulated light
curves.
Figure \ref{fig:compare_acf} shows ACF-measured versus true rotation
periods, with the $2n$ and $\frac{1}{2}n$ harmonic lines as dashed lines.
The MRD of the ACF-recovered periods is \percentacfMAD\ \% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
The injected and recovered rotation periods generally agree well,
though with a few drastic over- or underestimates.
Additionally, of the points clustered close to the 1:1 line,
more fall slightly below than above it, \ie\
rotation periods are preferentially underestimated.
This stems from how peak positions in the ACF are measured.
ACFs of stellar light curves have similar functional forms to the QP kernel
function in Equation \ref{eq:QP}: periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left
(towards shorter periods), because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect; however, standard practice
simply measures the peak position without taking it into account.
We adopt this standard practice here to faithfully compare
our new method to that used in the literature.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_acf.pdf}
\caption[ACF results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using the ACF technique.
    Several light curves have over-estimated rotation periods and some
    are drastically underestimated.}
\label{fig:compare_acf}
\end{center}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_pgram.pdf}
\caption[LS periodogram results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using a LS periodogram technique.
In many cases a large peak at a long period was present in the periodogram,
    producing a significant over-estimate of the period.
    }
\label{fig:pgram_compare}
\end{center}
\end{figure*}

>>>>

\newcommand{\dfmcomment}[1]{{\color{red}#1}}
\newcommand{\tdmcomment}[1]{{\color{blue}#1}}
\newcommand{\vrcomment}[1]{{\color{magenta}#1}}
\newcommand{\racomment}[1]{{\color{green}#1}}

@@

\begin{abstract}
The light curves of spotted, rotating stars often vary in a non-sinusoidal and
quasi-periodic fashion---spots move on the stellar surface and have finite
lifetimes, causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid therefore cannot accurately model a
rotationally modulated stellar light curve.
Physical models of the stellar surface also have many drawbacks preventing
    effective inference, such as highly degenerate or high-dimensional
    parameter spaces.
% \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In this work, we introduce an appropriate {\it effective} model: a Gaussian
    Process with a quasi-periodic covariance kernel function.
Using this highly flexible model, we are able to sample from the posterior
    Probability Density Function (PDF) of the periodic parameter,
    marginalising over the other kernel hyperparameters using a Markov Chain
    Monte Carlo approach.
To test the effectiveness of this method, we infer rotation periods from
\naigrain\ simulated stellar light curves, demonstrating that our results
    are more accurate than both a sine-fitting periodogram and an
    autocorrelation function method.
We also demonstrate that it works effectively on real data, by
inferring rotation periods for \nkois\ \Kepler\ stars.
Because this method delivers posterior PDFs it will enable hierarchical
    studies involving stellar rotation, particularly those involving
    population modelling, such as inferring stellar ages, obliquities in
    exoplanet systems, or characterising star-planet interactions.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

Standard methods to measure rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[\eg][]{Reinhold2013}, Auto-Correlation Functions (ACFs),
\citep[\eg][]{Mcquillan13b} and wavelet transforms \citep[\eg][]{Garcia2014}.
The precisions of the LS periodogram and wavelet methods are limited by the
suitability of the model choice: a sinusoid for the LS periodogram,
and a choice of mother wavelet, assumed to describe the data
over a range of transpositions \citep[see, \eg][]{Carter2010},
for the wavelet method.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, as long as the signal is approximately periodic the ACF will
display a peak at the rotation period, no matter its shape.
A drawback of the ACF method, however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.}, which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for most ground-based measurements, such
as the future Large Synoptic Survey Telescope (\LSST).
An ACF is also an operation performed on the data rather than a generative
model of the data, and so is not inherently probabilistic.
This means that the effects of the observational uncertainties cannot be
formally propagated to constraints on the rotation period.
% Many rotation periods in the literature have been inferred by measuring the
% position of the first peak in an ACF, however this approach can be dangerous.
% The exponential decay in correlation can shift the peak position short-wards
% of its true value, leading to an underestimate of the period.
% We return to this point in section \textsection \ref{sec:discussion}.

In this work, we introduce an {\it effective} model for rotationally
modulated stellar light curves
which captures the salient behaviour but is not
physically motivated---although some parameters may indeed be
{\it interpreted} as physical ones.
An ideal effective model should have a small number of non-degenerate parameters
and be flexible enough to perfectly capture non-sinusoidal and QP behaviour.
A Gaussian process (GP) model fulfills these requirements. We thus use a GP
as the generative model at the core of a method to probabilistically
infer accurate and precise stellar rotation periods.  This enables us to
estimate the posterior PDF of the rotation period, thereby producing a
justified estimate of its uncertainty.
% while simultaneously marginalizing over
% parameters which model correlated noise.

@@

GP models parameterise the covariance between data points by means of a
kernel function.
As a qualitative demonstration, we present the time-series in Figure
\ref{fig:GP_example}: the \kepler\ light curve of KIC \kepexample.
This is a relatively active star that rotates once every $\sim$
\kepexampleperiod\ days, with stochastic variability typical of \kepler\ FGK
stars.
Clearly, data points in this light curve are correlated.
Points close together in time are tightly correlated, and points more
widely separated are loosely correlated.
A GP models this variation in correlation as a function of the separation
between data points; that is, it models the {\it covariance structure} rather
than the data directly.
This lends GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas modelling the time series itself
might require much more complexity.
Figure \ref{fig:GP_example} demonstrates how a GP model fits the light curve of
KIC \kepexample.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

A range of kernel functions can describe stellar variability.
For example, the most commonly used kernel function, the `Squared
Exponential' (SE), defined as follows, could adequately fit the
KIC \kepexample\ light curve:
\begin{equation}
\label{eq:SE}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
Here $A>0$ is the amplitude of covariance, $l$ is the length scale of
exponential decay, and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple, with
just two parameters, $A$ and $l$.
If $l$ is large, two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable, making it possible to model a data set
and its derivatives simultaneously.
% \citep[see \eg][]{Rajpaul2015}
However, The SE kernel function does not well describe the covariance
in stellar light curves, nor is it {\it useful} for the problem of
rotation period inference because it does not capture periodic behaviour.
Inferring rotation periods thus requires a periodic kernel
function.
For this reason, we use the `Quasi-Periodic' kernel.
\citet{Rasmussen2005} model QP variability in CO$_2$ concentration on the
summit of the Mauna Loa volcano in Hawaii \citep[data from][]{Keeling2004}
using a kernel which is the product of a periodic and a SE kernel: the QP
kernel.
This kernel is defined as
\begin{equation}
\label{eq:QP}
k_{i,j} = A \exp \left[-\frac{(x_i - x_j)^2}{2l^2} -
    \Gamma^2 \sin^2\left(\frac{\pi(x_i - x_j)}{P}\right) \right] + \sigma^2
    \delta_{ij}.
\end{equation}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star, and $\Gamma$
controls the amplitude of the $\sin^2$ term.
If $\Gamma$ is very large, only points almost exactly one period away are
tightly correlated and points that are slightly more or less than one period
away are very loosely correlated.
If $\Gamma$ is small, points separated by one period are tightly
correlated, and points separated by slightly more or less are still highly
correlated, although less so.
In other words, large values of $\Gamma$ lead to periodic variations with
increasingly complex harmonic content.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while also allowing
points separated by half a period to be weakly correlated.
We also use an extra parameter, $\sigma$, which is an additional white noise
term added to the diagonal elements of the covariance matrix.
This can be interpreted to represent underestimation of observational
uncertainties~---~if the uncertainties reported on the data are too small, it
will be non-zero~---~or it can capture any remaining ``jitter'' or residuals
that are not captured by the effective GP model.
We use this QP kernel function to produce the GP model
that fits the \Kepler\ light curve in in Figure \ref{fig:GP_example}.
There are many ways to construct a QP kernel function, involving a range of
choices for both the periodic and a-periodic components of the model.
The kernel function presented above reproduces the behaviour of stellar light
curves but it is not necessarily the only model choice that can do so.
We do not attempt to test any other choices in this paper, noting only that
this kernel function provides an adequate fit to the data.
We leave formal comparisons with other kernel function choices to a future
publication.

To infer a stellar rotation period $P$ from a light curve, we fit this
QP-kernel GP model to the data.  As with any model-fitting exercise, the
likelihood
of the model could be maximised to find the maximum-likelihood value for $P$.
In this study, however, we explore the full posterior PDFs using a Markov
Chain Monte Carlo (MCMC) procedure.  While
this approach comes at a computational cost, such posterior exploration
importantly provides a justified uncertainty estimate.

@@

This paper is laid out as follows.
The GP method is described in \textsection \ref{sec:method}.
Its performance is demonstrated and compared with literature methods in
in \textsection \ref{sec:perf_and_comp}.
In \textsection \ref{sec:kepler} we apply our method to real \kepler\ data,
and the results are discussed in \textsection \ref{sec:discussion}.

@@

The flexibility of this GP model allows for posterior multimodality and
``over-fitting''--like behavior.
For example, if $l$ is small, the non-periodic factor in the covariance
kernel may dominate, allowing for a good fit to the data without
requiring any periodic covariance structure---even if clear periodic
structure is present.
Additionally, if $\Gamma$ is too large, the GP model becomes extremely
flexible and can fit the data without varying the period.
Managing this flexibility to reliably retrieve the period parameter requires
imposing informative priors on the non-period GP parameters.
% While there are no strictly {\it a priori} reasons to believe
% that any of the hyperparameters should take values in any specific range,
% we have found in this work that reasonable priors for any particular
% dataset can be determined by experimentation.
% We note this is a somewhat subjective art, with a goal of
% not allowing the GP to become so flexible that it does not require
% periodicity to fit the data even in the presence of clear periodicity.
In particular, we find it necessary to avoid large values of
$A$ and $\Gamma$, and small values of $l$; though the exact details of what
works well may differ among datasets.
See \textsection \ref{sec:discussion} for a discussion of more physically
motivated priors on the non-period parameters.

@@

% While using an uninformed prior on period (e.g., log-flat) often suffices,
% we also develop a method to construct an informed period prior, based on the
% autocorrelation function (ACF).
We use an informed period prior, based on the autocorrelation function (ACF)
of the light curve.
The ACF has proven to be very useful for measuring stellar rotation periods
\citep{Mcquillan2012, Mcquillan13b, Mcquillan2014}; however, the
method has several shortcomings,
most notably the inability to deliver uncertainties, but also
the necessity of several heuristic choices,
such as a timescale on which to smooth the ACF,
how to define a peak, whether the first or second peak
gets selected, and what constitutes a secure detection.
While this paper presents a rotation period inference method
that avoids these shortcomings,
it seems prudent to still use information available from the ACF.
We thus use the ACF to define a prior on period,
which can help the posterior sampling converge on the true period.

@@

\item{We assign a quality metric $Q_i$ to each of these candidate
periods, as follows.  First, we model the ACF as a
damped oscillator with fixed period $T_i$:
\begin{equation}
y = A e^{-t/\tau} \cos{\frac{2\pi t}{T_i} },
\end{equation}
        \racomment{where and $t$ is the lag time, (some missing text here.)}
and find the best-fitting parameters $A_i$ and $\tau_i$ by a non-linear
least squares minimization procedure.  We then define the
following heuristic quality metric:
\begin{equation}
\label{eq:quality}
Q_i = \left(\frac{\tau_i}{T_i}\right) \left(\frac{N_i h_i}{R_i}\right),
\end{equation}
where $h_i$ is the height of the ACF peak at $T_i$,
$N_i$ is the length of the lag vector in the ACF (directly proportional
to the maximum allowed period $P_i$),
and $R_i$ is the sum of squared residuals between the
damped oscillator model and the actual ACF data.  The idea behind this
quality metric is to give a candidate period a higher score if

@@

We initialise 500 walkers with random samples from the prior and use a
weighted mixture of the KDE, DE, and snooker proposals with weights of 0.4,
0.4, and 0.2 respectively.
We run \vrcomment{50} steps of the sampler at a time, checking for convergence
after each iteration, up to a maximum of \vrcomment{50} iterations.
We declare convergence if the total effective chain length is at least
8$\times$ the maximum autocorrelation time.
When convergence is achieved, we discard the first two autocorrelation lengths
in the chain as a burn-in, and randomly choose 5000 samples as representative
of the posterior.
This fitting process takes several hours for a typical simulated light curve,
though in some cases it can take 12 hours or longer to converge.

@@

In order to benchmark this new rotation period recovery method, we apply
it to a set of simulated light curves
and compare to the performance of established literature methods.
Section \ref{sec:simulations} describes the simulated data we use;
Section \ref{sec:performance} demonstrates the performance of the
QP-GP method; and Section \ref{sec:comparison} compares it to the performance
of the Lomb-Scargle periodogram and autocorrelation function methods.

@@

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarise our results compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
To assess the performance of the QP-GP and other period recovery
methods, we compute Median Absolute Deviations (MADs) of the results, relative
to the input periods.
We also compute the Median Relative Deviation (MRD), as a percentage.
These metrics are presented for the three different methods tested in this
paper in table \ref{tab:MADs}.
The informative and uninformative prior versions of the GP method have MRDs of
\percentgpMAD \% and \percentgpMADnp \% respectively.
The marginal posterior distributions of the QP kernel hyperparameters, for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{figures/compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{figures/comparison_acfprior_02_13.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
Since the posterior PDFs of rotation periods are often non-Gaussian,
    the points plotted here are maximum {\it a-posteriori} results.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
The ACF-informed prior on rotation period used to generate these results is
    described in \textsection \ref{sec:GP_prior}.
    }
\label{fig:compare_mcmc_acfprior}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/comparison_noprior_02_13.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique with an uninformative prior.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
Since the posterior PDFs of rotation periods are often non-Gaussian,
    the points plotted here are maximum {\it a-posteriori} results.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
An uninformative prior, flat in the natural log of the rotation period between
    0.5 and 100 days was used to generate these results.
    }
\label{fig:compare_mcmc_noprior}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/corner_plot.pdf}
\caption{Marginal posterior PDFs of the QP GP model parameters, fit to the
    simulated light curve in figure \ref{fig:demo_lc}.
    The blue line in the period panel shows the injected period.
    This figure was made using \texttt{corner.py} \citep{Corner}.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

@@

We apply the ACF method to the sample of \naigrain\ simulated light
curves.
Figure \ref{fig:compare_acf} shows ACF-measured versus true rotation
periods, with the $2n$ and $\frac{1}{2}n$ harmonic lines as dashed lines.
The MRD of the ACF-recovered periods is \percentacfMAD\% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
The injected and recovered rotation periods generally agree well,
though with a few drastic over- or underestimates.
Additionally, of the points clustered close to the 1:1 line,
more fall slightly below than above it, \ie\
rotation periods are preferentially underestimated.
This stems from peak position measurements in the ACF method.
ACFs of stellar light curves have similar functional forms to the QP kernel
function in Equation \ref{eq:QP}: periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left
(towards shorter periods), because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect; however, standard practice
simply measures the peak position without taking it into account.
We adopt this standard practice here to faithfully compare
our new method to that used in the literature.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_acf.pdf}
\caption[ACF results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using the ACF technique.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
    Several light curves have over-estimated rotation periods and some
    are drastically underestimated.}
\label{fig:compare_acf}
\end{center}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_pgram.pdf}
\caption[LS periodogram results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using a LS periodogram technique.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
In many cases a large peak at a long period was present in the periodogram,
    producing a significant over-estimate of the period.
    }
\label{fig:pgram_compare}
\end{center}
\end{figure*}
